---
title: "What is Due & When?"
subtitle: Fall 2019
output:
  html_document:
    df_print: paged
  pdf_document: default
---

You are expected to come to each class meeting prepared, which means that you will have done the readings and DataCamp assignments that are due for that date *before* coming to class. 

## Readings

Readings will primarily be from two free, open-source, completely online textbooks:

1. [**MODERN DIVE into Data with R (MD)**](http://moderndive.netlify.com) 
1. [**R for Data Science (R4DS)**](http://r4ds.had.co.nz) 

These books will be supplemented by blog posts, YouTube videos, and several open-access articles, including from:

- A special PEERJ issue called ["Practical Data Science for Stats"](https://peerj.com/collections/50-practicaldatascistats/), and
- [Nature: Points of Significance](https://www.nature.com/collections/qghhqm/pointsofsignificance)


## Ramping Up
The following DataCamp courses, tutorials, and readings are optional, but will help you come up to speed with the skills you'll need to complete this class. Unless you have strong, recent experience with R, RStudio, RMarkdown and the TidyVerse, I strongly suggest you complete, or at least review, them.

- [Markdown Tutorial](https://www.markdowntutorial.com)

DataCamp

- [Intro the to Tidyverse](https://www.datacamp.com/courses/introduction-to-the-tidyverse)
- [Working with data in the Tidyverse (long)](https://www.datacamp.com/courses/working-with-data-in-the-tidyverse)

ModernDive

- [MD Ch 1: Getting Started](https://moderndive.com/1-getting-started.html)
- [MD Ch 2: Data Visualization](https://moderndive.com/2-viz.html)	
- [MD Ch 3: Data Wrangling](https://moderndive.com/3-wrangling.html)	
- [MD Ch 4: Tidy Data](https://moderndive.com/4-tidy.html)	

R4DS (The below are chapter suggestions. The entire site serves as a good reference.)

- [R4DS: Workflow Projects](http://r4ds.had.co.nz/workflow-projects.html) 
- [R4DS: R Markdown](http://r4ds.had.co.nz/r-markdown.html)	
- [R4DS: Data Visualisation](http://r4ds.had.co.nz/data-visualisation.html)	
- [R4DS: Data Transformation](http://r4ds.had.co.nz/transform.html)	
- [R4DS: Tidy Data](http://r4ds.had.co.nz/tidy-data.html)	



```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(DT)
due <- read_csv("due.csv") %>%
  mutate(Date = paste(Month, Day)) %>%
  mutate(`DataCamp` = ifelse(material == "-" | is.na(material_link),
    material,
    paste0("<a href='", material_link, "'>", material, "</a>"))) %>% 
  mutate(`class_lab` = ifelse(class_lab == "-" | is.na(cl_link),
    class_lab,
    paste0("<a href='", cl_link, "'>", class_lab, "</a>"))) %>% 
  mutate(hw_due = ifelse(hw == "-" | is.na(hw) | is.na(hw_link),
    hw,
    paste0("<a href='", hw_link, "'>", hw, "</a>"))) %>% 
  mutate(i_lab_due = ifelse(i_lab == "-" | is.na(i_lab) | is.na(i_lab_link),
    i_lab,
    paste0("<a href='", i_lab_link, "'>", i_lab, "</a>"))) %>% 
  mutate(px_due = ifelse(px == "-" | is.na(px) | is.na(px_link),
    px,
    paste0("<a href='", px_link, "'>", px, "</a>"))) %>% 
  # select(Weekday, Date, `CM / Block`, `Readings Due` = Reading, 
  #        `DataCamp Due`, `Lab Due`, `Integrative Lab Due` = big_lab_due)
  select(Weekday, Date, `CM / Block`, `Readings` = Reading,
         DataCamp, `Homework & Midterm` = hw_due, `In-class Lab` = class_lab,`I_Lab` = i_lab_due, `Final Project` = px_due)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(knitr::is_html_output()){
  due %>% 
    datatable(rownames = FALSE,
              class = 'cell-border stripe',
              # filter = list(position = 'top'),
              options = list(pageLength = nrow(due), 
                             autoWidth = TRUE,
                             columnDefs = list(list(width = '280px', targets = c(3)),
                                               list(width = '30px',  targets = c(0)) ),
                             bInfo = FALSE,
                             paging = FALSE),
              escape = FALSE)
} else {
  knitr::kable(due, format = "latex", booktabs = TRUE, longtable = TRUE)
}
```

```{r echo=FALSE, eval=FALSE}
# how to get links into the pdf cleanly - modify the above to make sense in pdf format
urls <- rep("https://stackoverflow.com/", 10)
# use paste0 to compose markdown-formatted hyperlinks
mtcars$mpg <- paste0("[", mtcars$mpg, "](", urls, ")")
# print the table, with hyperlinked text
knitr::kable(head(mtcars))
```
