---
title: "Math 530/630 ncp"
subtitle: "The non-centrality parameter of the *t* distribution"
output:
  xaringan::moon_reader:
    css: ["default", "css/ohsu.css", "css/ohsu-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-lakeside-light
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
---

```{r setup_new, include=FALSE}
# leave this chunk alone
options(htmltools.dir.version = FALSE)
options(knitr.table.format = "html") 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300, out.width = '60%', cache = FALSE)

# knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
#   comment = NA, dpi = 300,
#   fig.align = "center", cache = FALSE)

```

``` {r libraries, include = FALSE} 
library(tidyverse)
library(infer)
library(moderndive)
```


# The non-centrality parameter of the *t* distribution

In an earlier class, I alluded to the fact that the *t*-distribution has an another parameter in addition to the degrees of freedom- a non-centrality parameter, $\delta$. 

> When $H_0$ is true, $\delta=0$.

> When $H_0$ is false, $\delta\neq0$.

Why is this? Let's look at the formula:

$$\delta=\sqrt{\frac{n_1n_2}{n_1+n_2}}(\frac{\mu_1-\mu_2}{SE})$$

As you can see, if $\mu_1-\mu_2$=0 then $\delta$=0.

---


# The *t*-distribution... 

.pull-left[

> When the null is true

```{r echo=FALSE}
# Script 12.4 - The Normal distribution and several t-distributions
# Clearing out the workspace
rm(list = ls())
# Setting the values for the abscissa
tval <- pretty(c(-3.25, 3.25), 1000)
# Find the ordinates for the normal distribution
ordn <- dnorm(tval)
# Drawing the normal distribution
plot(tval, ordn, type = "l", lty = 1, lwd = 3,
  ylab = "Ordinate", xlab = "Value")
# Setting df for three different t-distributions
nu <- c(1, 3, 10)
# Drawing the three different t-distributions
lines(tval, dt(tval, nu[1]), type = "l" ,lty = 2, lwd = 2)
lines(tval, dt(tval, nu[2]), type = "l", lty = 3, lwd = 2)
lines(tval, dt(tval, nu[3]), type = "l", lty = 5, lwd = 2)
legend(-3,.40, c("Normal", "t, df=1", "t, df=3", "t, df=10"),
  lwd = c(3,2,2,2), lty = c(1,2,3,5))
```
]

.pull-right[

> When the null is false

```{r echo=FALSE}
# Script 12.4 - The Normal distribution and several t-distributions
# Clearing out the workspace
rm(list = ls())
# Setting the values for the abscissa
tval <- pretty(c(-3.25, 3.25), 1000)
# Find the ordinates for the normal distribution
ordn <- dnorm(tval)
# Drawing the normal distribution
plot(tval, ordn, type = "l", lty = 1, lwd = 3,
  ylab = "Ordinate", xlab = "Value")
# Setting df for three different t-distributions
nu <- c(9, 9, 9)
delta <- c(.5, 1, 1.5)
# Drawing the three different t-distributions
lines(tval, dt(tval, nu[1], delta[1]), type = "l" ,lty = 2, lwd = 2)
lines(tval, dt(tval, nu[2], delta[2]), type = "l", lty = 3, lwd = 2)
lines(tval, dt(tval, nu[3], delta[3]), type = "l", lty = 5, lwd = 2)
legend(-3,.40, c("Normal", "t, delta=.5", "t, delta=1", "t, delta=1.5"),
  lwd = c(3,2,2,2), lty = c(1,2,3,5))
```
]

---

# Non-central *t*-distributions... 

.pull-left[

> Degrees of freedom=9

```{r echo=FALSE}
nu <- c(9, 9, 9, 9)
delta <- c(0, .5, 1, 1.5)
plot(tval, dt(tval, nu[1], delta[1]), type = "l" ,lty = 1, lwd = 3,
  ylab = "Ordinate", xlab = "Value", col="#d7191c")
# Drawing the three different t-distributions
lines(tval, dt(tval, nu[2], delta[2]), type = "l" ,lty = 2, lwd = 2, col="#fdae61")
lines(tval, dt(tval, nu[3], delta[3]), type = "l", lty = 3, lwd = 2, col="#abdda4")
lines(tval, dt(tval, nu[4], delta[4]), type = "l", lty = 5, lwd = 2, col="#2b83ba")
legend(-3,.40, c("t, delta=0", "t, delta=.5", "t, delta=1", "t, delta=1.5"),
  lwd = c(3,2,2,2), lty = c(1,2,3,5), col=c("#d7191c", "#fdae61","#abdda4", "#2b83ba"))
```
]

.pull-right[

> $\delta$=5

```{r echo=FALSE}
# Setting the values for the abscissa
ntval <- pretty(c(0,12), 1000)
nu <- c(10, 20, 30, 40)
delta <- c(5,5,5,5)
plot(ntval, dt(ntval, nu[1], delta[1]), type = "l" ,lty = 1, lwd = 3,
  ylab = "Ordinate", xlab = "Value", xlim=c(0, 12), ylim=c(0, .45), col="#d7191c")
# Drawing the three different t-distributions
lines(ntval, dt(ntval, nu[2], delta[2]), type = "l" ,lty = 2, lwd = 2, col="#fdae61")
lines(ntval, dt(ntval, nu[3], delta[3]), type = "l", lty = 3, lwd = 2, col="#abdda4")
lines(ntval, dt(ntval, nu[4], delta[4]), type = "l", lty = 5, lwd = 2, col="#2b83ba")
legend(0,.45, c("df=10", "df=20", "df=30", "df=40"),
  lwd = c(3,2,2,2), lty = c(1,2,3,5), col=c("#d7191c", "#fdae61","#abdda4", "#2b83ba"))
```
]

---

# Using non-central *t*-distributions... 

```{r}
help(TDist)
pt(-1, 20, 0)
pt(-1, 20, 5)
```

Recommended:
```{r eval=FALSE}
power.t.test()
```

