---
title: "Final Project"
author: "MATH 530/630"
subtitle: 'Part 0: Pick Groups and a Paper'
output:
  html_document: null
  pdf_document: default
  code_folding: hide
---


## Overview


The goal of this project piece is to choose your final project groups and to pick a paper you'll use for your final project. No overlap will be allowed - that is each paper can be used by only one group. Because of this, each group should submit their top 3 choices, and we'll do what we can to accomodate. 


NOTE: If the link below does not have the full pdf, you can find the papers using the [OHSU Library](https://www.ohsu.edu/xd/education/library/).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readxl) #reading in excel files
library(dplyr) #coalesce
library(tidyr) #spread
library(knitr) #kable
library(kableExtra) #add_header
```


```{r include=FALSE}
data_source <- here::here("project", "P0-pick-papers", "PaperChoices2019.xlsx")
sheets <- excel_sheets(data_source)
paper_data <- read_excel(data_source,sheets[1])
paper_data_cleaned <- paper_data %>%  
    mutate(Study = paste0("<a href='", Paper, "'>", Study, "</a>")) %>% 
    mutate(RepoLink = paste0("<a href='", RepoLink, "'>Data Repo Link</a>")) %>% 
  select(PaperNum, Study, RepoLink)
  
groups <- read_excel(data_source,sheets[2])
choices <- read_excel(data_source,sheets[3])
comb_choices <- left_join(paper_data_cleaned, choices)
comb_choices<- replace_na(comb_choices,list(Groups=0, Preference=0))

```



```{r include=FALSE}
# palce holder to produce a paper list when there are no groups formed
options(knitr.kable.NA = '')
paper_data_cleaned %>% 
  kable(format="html", escape=FALSE) %>% 
  kable_styling()

```


```{r eval=TRUE}
# Picks so far.... - set eval = TRUE when there's data in the spreadsheet

comb_table <- comb_choices %>%
  group_by("Paper" = PaperNum, Study, RepoLink, Preference) %>%
  summarise(n=n()) %>%
  spread(Preference, n)%>%
  select(-`0`)

#Should be tidied...
options(knitr.kable.NA = '')
kab_table <- kable(comb_table, format="html", escape = FALSE)
add_header_above(kab_table, c(" " = 3,"Preference (counts)" = 3)) %>% 
    kable_styling(bootstrap_options = c("hover", "condensed"))
```


```{r eval=TRUE}
# Groups - set eval to TRUE when there is data in the spreadsheet
options(knitr.kable.NA = '')
groups %>% 
  kable(align = c('l','c')) %>% 
  collapse_rows(columns = 1) %>% 
  kable_styling()

```

