---
title: 'Final Project'
author: "MATH 530/630"
subtitle: 'Part 0: Pick Groups and a Paper'
output:
  html_document:
    highlight: textmate
    smart: no
    theme: flatly
---


## Overview

Due Tuesday October 2 by 2pm (submit on [Sakai](https://sakai.ohsu.edu)).

The goal of this project piece is to choose your final project groups and to pick a paper you'll use for your final project. No overlap will be allowed - that is each paper can be used by only one group. Because of this, each group should submit their top 3 choices, and we'll do what we can to accomodate. 


NOTE: If the link below does not have the full pdf, you can find the papers using the [OHSU Library](https://www.ohsu.edu/xd/education/library/) .

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(DT)
due <- read_csv("PaperList.csv") %>%
    mutate(Paper = paste0("<a href='", Paper, "'>", Paper, "</a>")) %>% 
    mutate(RepoLink = paste0("<a href='", RepoLink, "'>", RepoLink, "</a>")) %>% 
  select(Study, Paper, RepoLink)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
if(knitr:::is_html_output()){
  due %>% 
    datatable(rownames = FALSE,
              class = 'cell-border stripe',
              # filter = list(position = 'top'),
              options = list(pageLength = nrow(due), 
                             autoWidth = TRUE,
                             bInfo = FALSE,
                             paging = FALSE),
              escape = FALSE)
} else {
  knitr::kable(due, booktabs = TRUE, longtable = TRUE)
}
```



<!-- The goal of this project piece is to look at and explore your dataset for your final project. You will use the [Quartz Bad Data Guide](https://github.com/Quartz/bad-data-guide/) to help guide your data quality review. Please use this  [`.csv`](quartz-bad-data-checklist.csv) file to indicate which, if any variable(s), in your dataset has any of the issues listed. If you indicate a variable with an issue, please include a brief paragraph (2-3 sentences tops) describing the issue further.  -->